---
title: "02_models_integration.R"
author: "BRW"
date: "2025-10-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Integrating SAFEKAW submodels
- Goal of this file is to integrate submodels (components) into one output via model wrappers on HISTORICAL DATA, baseline for optimization

Setup (paths + packages + model wrappers + load CSVs)
```{r}
library(lme4)
library(performance)
library(MuMIn)
library(here)


getwd()
source(here("code","paths+packages.R"))          #loads path & packages (from original files)
source(here("code/optimization", "01.5_model_wrappers.R"))   # loads model wrappers

yrs_common <- 2006:2023


df_combined <- read_csv(here("data", "common_inputs_county.csv")) 
common_input_basin <- read_csv(here( "data","common_inputs_basin.csv")) %>% mutate(Management_FertilizerUse_kgm2 = Management_FertilizerUse_kg / `LandCover_m2_all cult`)  |> subset(Year %in% yrs_common) 

df_county_areas  <- read_csv(here("data","county_areas.csv"))

```

# Irrigation model
 2a) Validation on training subset
```{r}
irr_val <- run_irrigation_stat_existing(df_combined)
print(irr_val$diagnostics)

r2_df_irr <- round(as.numeric(irr_val$diagnostics[3]), 2)

r2_df_irr

summary(irr_val$model)
coef_irr <- summary(irr_val$model)$coefficients
print(coef_irr)


p_irr_val <- irr_val$data %>%
  ggplot(aes(x = irrigation_WaterUse_m, y = irr_pred_m)) + #precip = negative, tmax_C_avg = postive
  geom_point(aes(color = Crop), alpha = 0.7) +
   scale_color_manual( values =c("#EEC900", "#548B54")) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1, linetype = "dashed") +
  labs(x = "Observed irrigation depth (m)",
       y = "Predicted irrigation depth (m)",
       title = "Observed vs Predicted Irrigation") +
   geom_text(data = as.data.frame(r2_df_irr), aes(x = Inf, y = Inf, label =  paste0("R² = ",r2_df_irr)),
            inherit.aes = FALSE, hjust = 4.5, vjust = 2.5, size = 5) +
  theme_test(base_size = 15)
p_irr_val
ggsave(file.path("figures", "Irrigation_ObsVsPred_training.png"),
       width = 150, height = 120, units = "mm", dpi = 300)

#ggsave(file.path("..","Irrigation_ObsVsPred_training.png"), p_irr_val, width = 7, height = 5, dpi = 300)


p_irr_val_res<- irr_val$data %>%
  mutate(resid_m = irrigation_WaterUse_m - irr_pred_m) %>%
  ggplot(aes(x = Year, y = resid_m, color = Crop)) + #irr_pred_m, FIPS, precip_m, tmax_C_avg, Year
     scale_color_manual( values =c("#EEC900", "#548B54")) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Year", y = "Residual (obs - pred)",
       title = "Irrigation residuals vs prediction") +
  theme_test(base_size = 15)
p_irr_val_res
ggsave(file.path("figures", "Irrigation_TimevResiduals.png"),
       width = 150, height = 120, units = "mm", dpi = 300)

irr_val$data %>% summarise(bias = mean( irr_pred_m- irrigation_WaterUse_m))


#install.packages('Metrics')
library(Metrics)

actual <- irr_val$data$irrigation_WaterUse_m
predicted <- irr_val$data$irr_pred_m
bias(predicted, actual)




```


 
 2b) Predict for eligible crops only across all rows (Corn/Soy; NI -> 0)
```{r}
irr_out <- run_irrigation_predict_eligible(df_combined)
irr_aug <- irr_out$data_all_aug  
```


# Crop yield models 
Build modeling df with modeled irrigation -> totalWater
```{r}
df_yield <- read_csv(here("data", "CropYieldData_County.csv")) |>
  subset(Year %in% yrs_common) |>
  mutate(FIPS = as.character(FIPS))

df_crop <- irr_aug %>%
  filter(!is.na(Crop), Year %in% yrs_common) %>%
    mutate(FIPS = as.character(FIPS)) %>%
  left_join(df_yield, by = c("Year","FIPS","Crop")) %>%
  mutate(
   irrigation_WaterUse_m = if_else(WaterManagement %in% c("Rainfed","Non-Irrigated"), 0, irrigation_WaterUse_m),
    totalWater_m          = precip_m + irrigation_WaterUse_m
  ) %>%
  filter(!is.na(totalWater_m))

# Optional area threshold (if area_ha present)
if ("area_ha" %in% names(df_crop)) {
  area_ha_thres <- 64.75 * 10   # quarter-section * 10
  df_crop <- df_crop %>% filter(area_ha > area_ha_thres)
}

# Fit per-crop wrappers
crop_results <- run_crop_models(df_crop)
df_combo_aug <- crop_results$data
print(crop_results$fit_stats)

# Quick validation plots
p_wp <- ggplot(df_combo_aug, aes(x = totalWater_m, y = yield_kgHa_detrended)) +
  geom_point(aes(color = WaterManagement), alpha = 0.6) +
  facet_wrap(~ Crop, scales = "free_y") +
  labs(x = "Precipitation + Irrigation (m)", y = "Yield (kg/ha)",
       title = "Water–Productivity Curves (modeled irrigation)") +
  geom_line(aes(y = yield_kgHa_detrended_fit), color = "grey40", linewidth = 0.7, alpha = 0.6) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), color = "black") +
  theme_test(base_size = 12) + theme(legend.position = "bottom")
#ggsave(file.path("..", "Crop_WaterProductivity_byCrop.png"), p_wp, width = 8, height = 6, dpi = 300)
p_wp
ggsave(file.path("figures", "water_prod_curves.png"),
       width = 160, height = 150, units = "mm", dpi = 300)

r2_df <- crop_results$fit_stats %>%
  select(Crop, fit_R2_kcalHa) %>%
  mutate(label = paste0("R² = ", round(fit_R2_kcalHa, 2)))

p_predobs <- ggplot(df_combo_aug,
                    aes(x = yield_kgHa_detrended_fit, y = yield_kgHa_detrended)) +
  geom_abline(intercept = 0, slope = 1, color = "gray40") +
  geom_point(aes(color = WaterManagement), alpha = 0.6) +
  facet_wrap(~ Crop, scales = "free") +
  stat_smooth(method = "lm", color = "black") +
  labs(x = "Predicted Yield (kg/ha)", y = "Observed Yield (kg/ha)",
       title = "Predicted vs Observed Yield") +
  geom_text(data = r2_df, aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 4.1, vjust = 1.2, size = 3.5) +
  theme_test(base_size = 14) + 
  theme(legend.position = "none")
#ggsave(file.path("..", "Crop_PredVsObs_byCrop.png"), p_predobs, width = 8, height = 6, dpi = 300)
p_predobs
ggsave(file.path("figures", "Yields_ObsVsPred_training.png"),
       width = 160, height = 150, units = "mm", dpi = 300)



p_resids <- df_combo_aug %>% 
  mutate(resids = yield_kgHa_detrended -yield_kgHa_detrended_fit ) %>%
  
  ggplot(aes(x = precip_m, y = resids)) + #yield_kgHa_detrended_fit, FIPS, Year , GDD  precip_m
  #geom_abline(intercept = 0, slope = 1, color = "gray40") +
  geom_point(aes(color = WaterManagement), alpha = 0.6) +
  facet_wrap(~ Crop, scales = "free") +
  stat_smooth(method = "lm", color = "black") +
  labs(x = "Precip. (m)", y = "Residuals (obs - pred) (kg/ha)",
       title = "Precip. vs Residuals Yield") +
  #geom_text(data = r2_df, aes(x = Inf, y = Inf, label = label),
       #     inherit.aes = FALSE, hjust = 4.1, vjust = 1.2, size = 3.5) +
  theme_test(base_size = 14) + 
  theme(legend.position = "none")
#ggsave(file.path("..", "Crop_PredVsObs_byCrop.png"), p_predobs, width = 8, height = 6, dpi = 300)
p_resids
ggsave(file.path("figures", "Yields_resid_ppt.png"),
       width = 160, height = 150, units = "mm", dpi = 300)


```

 
  - more model checking
```{r}
mods       <- crop_results$models
crop_names2 <- sapply(mods, `[[`, "crop")
crop_idx2  <- which(crop_names2 == "Soybeans")
performance::r2(mods[[crop_idx2]]$fit_kcal)
performance::icc(mods[[crop_idx2]]$fit_kcal)

performance::check_normality(mods[[crop_idx2]]$fit_kcal)
performance::check_autocorrelation(mods[[crop_idx2]]$fit_kcal) 
performance::check_heteroscedasticity(mods[[crop_idx2]]$fit_kcal)


performance::check_collinearity(mods[[crop_idx2]]$fit_kcal)


crop_fit <- mods[[crop_idx2]]$fit_kcal
summary(crop_fit)   # check random effect variances, fixed-effect signs

# Residual vs fitted
plot(crop_fit)      

# QQ-plot of residuals
qqnorm(residuals(crop_fit)); qqline(residuals(crop_fit))



```
 
 
 #Water-quality model (basin)
```{r}
df_nitrate <- read_csv(file.path("data","RiverNitrateData_EKSRB.csv"), show_col_types = FALSE)

wq_input <- build_wq_input_from_basin(common_input_basin, irr_aug, nitrate_df = df_nitrate)


# Train only on complete rows
required <- c("NitrateFlux_kg","LandCover_m2_grassPastureHay",  "precip_percentileChange", ##"LandCover_m2_developed",
              "LandCover_m2_all cult","Climate_precip_m","Climate_Tmean_C",
             "Management_irrigation_m","Management_FertilizerUse_kg")

wq_train <- wq_input %>%  tidyr::drop_na(dplyr::all_of(required))

wq_results <- run_waterquality_model(wq_train)


df_wq_aug  <- wq_results$data
print(wq_results$performance)
r2_df_wq <- round(as.numeric(wq_results$performance[3]), 2)
r2_df_wq

coef_Nflux <- summary(wq_results$model)$coefficients
print(coef_Nflux)



summary(wq_results$model) 
car::vif(wq_results$model)



#how much unique variance does each term explain given all others?
drop1(wq_results$model, test = "F")

b <- coef(wq_results$model)
A_ref <- mean(wq_results$data$BasinArea_m2, na.rm = TRUE)

delta_frac_100km2 <- (100 * 1e6) / A_ref

delta_flux_100km2_cult <-
  b["cult_frac"] * delta_frac_100km2
delta_flux_1C          <- b["Climate_Tmean_C"] * 1
delta_flux_01m_precip  <- b["Climate_precip_m"] * 0.1
c(delta_flux_100km2_cult, delta_flux_1C, delta_flux_01m_precip)





# Simple validation plot if observed NitrateFlux_kg exists in common_input_basin
if ("NitrateFlux_kg" %in% names(df_wq_aug)) {
  p_wq <- df_wq_aug %>%
    select(Year, NitrateFlux_kg, NitrateFluxPredicted_kg) %>%
    pivot_longer(-Year, names_to = "Variable", values_to = "Value") %>%
    ggplot(aes(Year, Value, color = Variable)) +
    geom_point() + geom_line() +
    scale_color_manual(values = c("black","red")) +
    labs(y = "Nitrate Flux (kg/yr)", title = "Water Quality Model Validation") +
    theme_test(base_size = 12)
}

p_wq



ggplot(df_wq_aug, aes(x = NitrateFlux_kg, y = NitrateFluxPredicted_kg)) +
  #LandCover_m2_developed -, LandCover_m2_grassPastureHay ~-,  'LandCover_m2_all cult` ~,  Climate_precip_m +,   Climate_Tmean_C -, precip_percentileChange ~, Management_irrigation_m -, Management_FertilizerUse_kgm2 ~+
  geom_point(color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(x = "Observed Nitrate Flux (kg/yr)", 
       y = "Predicted Nitrate Flux (kg/yr)",
       title = "Observed vs Predicted Nitrate Flux") +
    geom_text(data = as.data.frame(r2_df_wq), aes(x = Inf, y = Inf, label =  paste0("R² = ",r2_df_wq)),
            inherit.aes = FALSE, hjust = 4.5, vjust = 2.5, size = 5) +
    theme_test(base_size = 14)
ggsave(file.path("figures", "Nflux_ObsVsPred_training.png"),
       width = 130, height = 120, units = "mm", dpi = 300)

df_wq_aug %>% 
 mutate(resid_m = NitrateFlux_kg - NitrateFluxPredicted_kg) %>%
  ggplot(aes(x = NitrateFluxPredicted_kg, y = resid_m)) + #irr_pred_m, FIPS, precip_m, tmax_C_avg, Year
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted N flux", y = "Residual (obs - pred)",
       title = "Residuals vs prediction") +
  theme_test(base_size = 15)

```
 
 
#Net Return (per-row, from modeled irrigation & fitted yield)
```{r}
universal_costs <- list(
  irr_cost_per_m3 = 0.029   # $/m³ (Kisekka et al., 2016)
)

crop_params <- tribble(
  ~Crop,           ~income_per_kg, ~direct_cost_per_kg, ~fixed_cost_per_kg, ~total_cost_per_kg, ~fert_kgHa,
  "Wheat",         0.2,           0.12,                0.05,               0.17,               90,
  "Corn",          0.18,           0.11,                0.05,               0.15,              250, 
  "Sorghum", 0.17,           0.09,                0.06,               0.15,                    110,
  "Soybeans",       0.37,           0.18,                0.14,               0.32,              55
) %>%
  
  mutate(
    net_return_per_kg = income_per_kg - total_cost_per_kg
  )

df_combo_NR <- compute_net_returns(df_combo_aug, crop_params, universal_costs)

```




 - aggregating at the two scales and visualizing
```{r}
# --- Basin per-year, per-crop (per-ha) summary from df_combo_NR ---
crop_basin_perha <- df_combo_NR %>%
  group_by(Year, Crop) %>%
  summarise(
    Crop_area_m2 = sum(area_m2), 
    Crop_area_ha = sum(area_ha),
    
    Yield_kgHa   = weighted.mean(yield_kgHa_detrended_fit, w = area_m2, na.rm = TRUE),
    Irrigation_m = weighted.mean(irr_depth_m_pred,         w = area_m2, na.rm = TRUE),

    Irrigation_m3 = sum(Irrigation_m3_total,     na.rm = TRUE),
    fert_kgcrop = sum(fertilizer_kgCrop, na.rm = TRUE),
    
    NetReturn_ha  = weighted.mean(NetReturn_dollar_ha, w = area_m2, na.rm = TRUE),
   
    Crop_prc =  mean(area_prc),

    .groups = "drop"
  )  %>%
  left_join(
    df_wq_aug %>% dplyr::select(Year, NitrateFluxPredicted_kg = NitrateFluxPredicted_kg),
    by = "Year"
  )



df_long_crop <- crop_basin_perha %>%
  select(Year, Crop, Yield_kgHa, NetReturn_ha, Irrigation_m, Crop_prc, fert_kgcrop) %>% #NitrateFluxPredicted_kg
  pivot_longer(
    cols = c(Yield_kgHa, NetReturn_ha, , Irrigation_m, Crop_prc, fert_kgcrop), #NitrateFluxPredicted_kg
    names_to = "Variable", values_to = "Value"
  ) %>%
  mutate(Variable = factor(
    Variable,
    levels = c("Yield_kgHa", "NetReturn_ha",  "Irrigation_m", "Crop_prc", "fert_kgcrop"), #"NitrateFluxPredicted_kg",
    labels = c("Yield (kg/ha)", "Net Return ($/ha)",  "Irrigation (m)", "Crop %", "Fertilizer (kg/yr)") #"Nitrate Flux (kg/yr)",
  ))

p_overview_crop <- ggplot(df_long_crop, aes(x = Year, y = Value, color = Crop)) +
  geom_line() + geom_point() +
  facet_wrap(~ Variable, scales = "free_y", ncol = 2) +
  labs(title = "Integrated Model Outputs Over Time (area norm x crop)", x = "Year", y = "Value") +
  theme_test(base_size = 12)


p_overview_crop
ggsave(file.path("figures", "Int_Basin_Crop.png"),
       width = 180, height = 120, units = "mm")


# Save table for downstream work
write_csv(crop_basin_perha, file.path("data", "int_crop_areanorm_annual.csv"))

```


```{r}
# --- Basin totals per year (join areas first) ---
area_by_crop <- df_combined %>%
  filter(!is.na(Crop)) %>%
  mutate(area_ha = coalesce(area_ha, area_m2/10000)) %>%
  group_by(Year, Crop) %>%
  summarise(area_ha_total = sum(area_ha, na.rm = TRUE),
            area_m2_total = sum(area_m2, na.rm = TRUE)
            , .groups = "drop")

basin_totals <- crop_basin_perha %>%
  left_join(area_by_crop, by = c("Year","Crop")) %>%
  mutate(
    Yield_total_kg      = area_ha_total * Yield_kgHa,
    NetReturn_total_usd = area_ha_total * NetReturn_ha
  ) %>%
  group_by(Year) %>%
  summarise(
    Yield_total_kg      = sum(Yield_total_kg,      na.rm = TRUE),
    NetReturn_total_usd = sum(NetReturn_total_usd, na.rm = TRUE),
    Irrigation_total_m3 = sum(Irrigation_m3, na.rm = TRUE),
    Area_m2 = sum(area_m2_total,  na.rm = TRUE), 
    Area_ha = sum(area_ha_total,  na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  left_join(
    df_wq_aug %>% dplyr::select(Year, NitrateFluxPredicted_kg = NitrateFluxPredicted_kg),
    by = "Year"
  )

write_csv(basin_totals, file.path("data", "int_basin_annual.csv"))


#viz
df_long_basin <- basin_totals %>%
  select(Year, Yield_total_kg, NetReturn_total_usd, NitrateFluxPredicted_kg, Irrigation_total_m3, Area_m2) %>%
  pivot_longer(
    cols = c(Yield_total_kg, NetReturn_total_usd, NitrateFluxPredicted_kg, Irrigation_total_m3, Area_m2),
    names_to = "Variable", values_to = "Value"
  ) %>%
  mutate(Variable = factor(
    Variable,
    levels = c("Yield_total_kg", "NetReturn_total_usd", "NitrateFluxPredicted_kg", "Irrigation_total_m3", "Area_m2"),
    labels = c("Yield (kg)", "Net Return ($)", "Nitrate Flux (kg)", "Irrigation (m³)", "Cult. Area (m2)")
  ))

p_overview_basin <- ggplot(df_long_basin, aes(x = Year, y = Value, color = Variable)) +
  geom_line() + geom_point() +
  facet_wrap(~ Variable, scales = "free_y", ncol = 2) +
  labs(title = "Integrated Model Outputs Over Time", x = "Year", y = "Value") +
  theme_test(base_size = 12)

p_overview_basin
ggsave(file.path("figures", "Int_Basin.png"),
       width = 180, height = 120, units = "mm")

```
  
- figure out current proportions of irr/raindfed ag to add constraint to opt
```{r}
irr_frac_YR <- 
  df_combo_NR %>% group_by(Year, WaterManagement) %>% summarise(area_m2 = sum(area_m2, na.rm = TRUE), .groups = "drop_last") %>% 
  mutate(total_area_allcrops = sum(area_m2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    frac_of_basin_cultivated = area_m2 / total_area_allcrops)


irr_frac_YR %>% ggplot(aes(Year, frac_of_basin_cultivated, color = WaterManagement)) +   #facet_wrap(~Crop) + 
  geom_line() + 
  theme_test(base_size = 16)

df_combo_NR %>% 
  distinct(Crop)

irr_frac_YR %>% filter(WaterManagement == "Irrigated") %>% summarise(mean = mean(frac_of_basin_cultivated))
```

