---
title: "R Notebook for SAFEKAW opt using NSGA-II"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

 - setting wd, loading files + sources  
```{r}
getwd()
library(here)
source(here(file.path("code","paths+packages.R")) )        #loads path & packages (from original files)
source(here(file.path("code", "01.5_model_wrappers.R")) )  # loads model wrappers

library("rPref")
library(colourpicker)
library(mco)


df_county_input <- read_csv(here(file.path("data", "common_inputs_county.csv")) )
df_basin_input_forWQ <- read_csv(here(file.path("data","common_inputs_basin.csv"))) %>% mutate(Management_FertilizerUse_kgm2 = Management_FertilizerUse_kg / `LandCover_m2_all cult`)
df_county_areas  <- read_csv(here(file.path("data","county_areas.csv")))
df_int_crop_areanorm <-  read_csv(here(file.path("data","int_crop_areanorm_annual.csv")) )  
df_int_basin <-  read_csv(here(file.path("data","int_basin_annual.csv"))   )


yrs_common <- 2006:2023


```

 - color palettes
```{r}

pal_landcover <- c(
  "Grass/Pasture/Hay" = "#C1FFC1", #"#9BCD9B",
  "Cultivated"       = "#8B8B00", #"#8B5742",
  "Developed"         = "darkgrey"
)

# crop types
pal_crops <- c(
  "Corn"    = "#EEC900",
  "Soybeans" = "#548B54",
  "Wheat"   = "#8B7355", ##CD6839,
  "Sorghum" =  "darkorange"       #"#8B475D"
)


# ggplot2 helper scales so you don't have to retype scale_*_manual every time
scale_fill_landcover <- function(...) {
  ggplot2::scale_fill_manual(values = pal_landcover, ...)
}
scale_color_landcover <- function(...) {
  ggplot2::scale_color_manual(values = pal_landcover, ...)
}

scale_fill_crops <- function(...) {
  ggplot2::scale_fill_manual(values = pal_crops, ...)
}
scale_color_crops <- function(...) {
  ggplot2::scale_color_manual(values = pal_crops, ...)
}

```

 - setting params for econ - can move to separate file if needed
```{r}
universal_costs <- list(
  irr_cost_per_m3 = 0.029   # $/m³ (Kisekka et al., 2016)
)

crop_params <- tribble(
  ~Crop,           ~income_per_kg, ~direct_cost_per_kg, ~fixed_cost_per_kg, ~total_cost_per_kg, ~ fert_kgha,
  "Wheat",         0.2,           0.12,                0.05,               0.17,               90,
  "Corn",          0.18,           0.11,                0.05,               0.15,              250, 
  "Sorghum", 0.17,           0.09,                0.06,               0.15,                    110,
  "Soybeans",       0.37,           0.18,                0.14,               0.32,              55
) %>%
  
  mutate(
    net_return_per_kg = income_per_kg - total_cost_per_kg
  )


mix_crops <- c("Corn","Soybeans","Sorghum","Wheat")

df_opt <- df_int_crop_areanorm %>%
  left_join(crop_params, by = "Crop")



```

 - getting hist land use fracs
```{r}
hist_crop_frac <- df_int_crop_areanorm %>%
  #filter(Crop %in% crop_params) %>%
  group_by(Year, Crop) %>%
  summarise(
    area_m2 = if ("Crop_area_m2" %in% names(.)) sum(Crop_area_m2, na.rm = TRUE)
              else if ("Crop_area_ha" %in% names(.)) sum(Crop_area_ha, na.rm = TRUE) * 1e4
              else NA_real_,
    .groups = "drop"
  ) %>%
  group_by(Year) %>%
  mutate(frac = area_m2 / sum(area_m2, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(Crop) %>%
  summarise(
    ymin = min(frac, na.rm = TRUE),
   # ymed = median(frac, na.rm = TRUE),
    ymax = max(frac, na.rm = TRUE),
    .groups = "drop"
  )

# Split out the bands for convenience
crop_min <- hist_crop_frac |>
  dplyr::select(Crop, ymin) |>
  tibble::deframe()

crop_max <- hist_crop_frac |>
  dplyr::select(Crop, ymax) |>
  tibble::deframe()
```
 

 - run wq model - same here
```{r}
df_nitrate <- read_csv(here(file.path("data","RiverNitrateData_EKSRB.csv")), show_col_types = FALSE)

df_basin_input_forWQ <- df_basin_input_forWQ %>%
  left_join(df_nitrate, by = "Year") %>%
  subset(Year %in% yrs_common)

# Percentile change as in training
ec <- ecdf(df_basin_input_forWQ$Climate_precip_m)
wq_context <- df_basin_input_forWQ %>%
  arrange(Year) %>%
  mutate(
    precip_percentile       = ec(Climate_precip_m),
    precip_percentileChange = tidyr::replace_na(precip_percentile - dplyr::lag(precip_percentile), 0)
  )

req_wq <- c(
  "NitrateFlux_kg","LandCover_m2_developed","LandCover_m2_grassPastureHay",
  "LandCover_m2_all cult","Climate_precip_m","Climate_Tmean_C",
  "precip_percentileChange","Management_irrigation_m","Management_FertilizerUse_kgm2"
)
wq_train <- wq_context %>% tidyr::drop_na(dplyr::all_of(req_wq))

# Use log1p model to enforce positivity at predict time
wq_model_log1p <- lm(
  log1p(NitrateFlux_kg) ~ LandCover_m2_developed + LandCover_m2_grassPastureHay + `LandCover_m2_all cult` +
    Climate_precip_m + Climate_Tmean_C + precip_percentileChange +
    Management_irrigation_m + Management_FertilizerUse_kgm2,
  data = wq_train
)
summary(wq_model_log1p)



```

 -  Baseline land cover per year (developed fixed; grass is the slack)
  - this will have to be projected forward in predictions to reflect previous land use patterns 
```{r}
lu_baseline <- df_basin_input_forWQ %>%
  select(
    Year,
    LC_cult_base  = `LandCover_m2_all cult`,
    LC_grass_base = LandCover_m2_grassPastureHay,
    LC_dev_base   = LandCover_m2_developed
  ) %>%
  mutate(LC_total = LC_cult_base + LC_grass_base + LC_dev_base)

```

 - IDK if this is needed
```{r}
# Allocate irrigated vs rainfed shares by crop
# crop_shares: named numeric vector (e.g. c(Corn=0.4, Soybeans=0.3, Sorghum=0.2, Wheat=0.1))
# theta_irrig: scalar in [0,1], overall irrigated fraction of cultivated area
# hist_mix: optional tibble/data.frame with columns: Crop, irrig_frac (historical irrig share per crop)
# mode: "preserve_hist_mix" or "uniform_by_crop"
allocate_irrigation <- function(crop_shares,
                                theta_irrig,
                                hist_mix = NULL,
                                mode = c("preserve_hist_mix", "uniform_by_crop")) {
  mode <- match.arg(mode)
  crop_shares <- crop_shares[!is.na(crop_shares)]
  theta_irrig <- max(0, min(1, theta_irrig))
  crop_shares[crop_shares < 0] <- 0
  
  if (sum(crop_shares) <= 0) {
    irrig  <- crop_shares * 0
    rainfd <- crop_shares * 0
    return(list(irrig = irrig, rainfed = rainfd))
  }
  
  crop_shares <- crop_shares / sum(crop_shares)
  
  if (mode == "uniform_by_crop" || is.null(hist_mix)) {
    irrig  <- crop_shares * theta_irrig
    rainfd <- crop_shares - irrig
  } else {
    if (!all(c("Crop", "irrig_frac") %in% names(hist_mix))) {
      irrig  <- crop_shares * theta_irrig
      rainfd <- crop_shares - irrig
    } else {
      hm <- hist_mix[match(names(crop_shares), hist_mix$Crop), , drop = FALSE]
      hist_frac <- hm$irrig_frac
      hist_frac[is.na(hist_frac)] <- 0
      hist_frac <- pmax(0, pmin(1, hist_frac))
      raw_irrig <- crop_shares * hist_frac
      if (sum(raw_irrig) <= 0) {
        irrig <- crop_shares * theta_irrig
      } else {
        scale_factor <- theta_irrig / sum(raw_irrig)
        irrig <- raw_irrig * scale_factor
        irrig <- pmin(irrig, crop_shares)
      }
      rainfd <- crop_shares - irrig
    }
  }
  
  list(irrig = irrig, rainfed = rainfd)
}

```


 - One-year scorer
  - this fx generates land use candidates (i.e., crop distributions within a single year), calculates water use, fert prod, NR per crop for that land use/crop type candidate for each year (n = 18).. 
```{r}
eval_candidate <- function(
  x, 
  Y,
  df_opt,
  wq_model_log1p,
  wq_context,
  lu_baseline,
  universal_costs,
  crop_params,
  baseline_irrig_frac,
  hist_mix = NULL  # tibble: Crop, irrig_frac_crop (or similar), can stay NULL for baseline
){
  PEN <- 1e12
  `%||%` <- function(a, b) if (is.null(a)) b else a
  
  # ---- 0) decision vector -> crop shares (+ leftover wheat) ---------
  corn <- x[1]; soy <- x[2]; sor <- x[3]
  wheat <- 1 - corn - soy - sor
  
  if (any(c(corn, soy, sor, wheat) < 0) || any(c(corn, soy, sor, wheat) > 1)) {
    return(c(PEN, PEN, PEN))
  }
  
  shares <- c(Corn = corn, Soybeans = soy, Sorghum = sor, Wheat = wheat)
  
  # ---- 1) baseline land cover for this year -------------------------
  lu_y <- lu_baseline[lu_baseline$Year == Y, ]
  if (nrow(lu_y) != 1L) return(c(PEN, PEN, PEN))
  
  LC_cult_base  <- lu_y$LC_cult_base
  LC_grass_base <- lu_y$LC_grass_base
  LC_dev_base   <- lu_y$LC_dev_base
  LC_total      <- lu_y$LC_total
  
  # No expansion: cultivated area fixed
  Cult_m2  <- LC_cult_base
  Grass_m2 <- LC_total - LC_dev_base - Cult_m2
  
  # ---- 2) per-ha rows (4 crops) ------------------------------------
  crops4 <- c("Corn","Soybeans","Sorghum","Wheat")
  
  perha <- df_opt %>%
    dplyr::filter(Year == Y, Crop %in% crops4) %>%
    tidyr::complete(
      Year = Y, Crop = crops4,
      fill = list(
        Yield_kgHa   = 0,
        Irrigation_m = 0
      )
    ) %>%
    # Attach econ + fert params (including fert_kgHa) from crop_params
    dplyr::select(-dplyr::any_of(names(crop_params)[names(crop_params) != "Crop"])) %>%
    dplyr::left_join(crop_params, by = "Crop") # %>%
    #dplyr::mutate(fert_kgha = dplyr::coalesce(.data$fert_kgha, .data$fert_kgha, 0))
  
  # ---- 3) irrigation allocation at baseline irrig fraction ---------
  alloc <- allocate_irrigation(
    crop_shares = shares,
    theta_irrig = baseline_irrig_frac,
    hist_mix    = hist_mix,           # NULL is fine for baseline
    mode        = "uniform_by_crop"   # i.e., same irrig fraction for all crops
  )
  
  area_irrig_m2 <- Cult_m2 * alloc$irrig
  area_rain_m2  <- Cult_m2 * alloc$rainfed
  area_irrig_ha <- area_irrig_m2 / 1e4
  area_rain_ha  <- area_rain_m2  / 1e4
  
  # ---- 4) build irrigated vs rainfed slices ------------------------
  perha_irrig <- perha %>%
    dplyr::mutate(
      area_m2 = as.numeric(area_irrig_m2[match(Crop, names(area_irrig_m2))]),
      area_ha = as.numeric(area_irrig_ha[match(Crop, names(area_irrig_ha))]),
      # 👉 use climate/model-based depth per ha from df_opt
      Irrigation_m_eff = Irrigation_m
    )
  
  perha_rain <- perha %>%
    dplyr::mutate(
      area_m2 = as.numeric(area_rain_m2[match(Crop, names(area_rain_m2))]),
      area_ha = as.numeric(area_rain_ha[match(Crop, names(area_rain_ha))]),
      Irrigation_m_eff = 0   # rainfed = no applied irrigation
    )
  
  perha2 <- dplyr::bind_rows(
    perha_irrig %>% dplyr::mutate(Management = "irrigated"),
    perha_rain  %>% dplyr::mutate(Management = "rainfed")
  )
  
  # ---- 5) per-crop totals using per-kg econ ------------------------
  perha2 <- perha2 %>%
    dplyr::mutate(
      revenue_total   = income_per_kg     * Yield_kgHa * area_ha,
      base_cost_total = total_cost_per_kg * Yield_kgHa * area_ha,
      irr_cost_total  = (Irrigation_m_eff * area_m2) *
                        (universal_costs$irr_cost_per_m3 %||% 0)
    )
  
  # For WQ accounting
  Fert_kg_total <- sum(perha2$fert_kgha * perha2$area_ha,       na.rm = TRUE)
  Irr_m3_total  <- sum(perha2$Irrigation_m_eff * perha2$area_m2, na.rm = TRUE)
  
  # Basin totals ($/yr)
  Revenue_total  <- sum(perha2$revenue_total,   na.rm = TRUE)
  BaseCost_total <- sum(perha2$base_cost_total, na.rm = TRUE)
  Irr_totalCost  <- sum(perha2$irr_cost_total,  na.rm = TRUE)
  
  # No land-use change cost in this baseline mode
  NR_total <- Revenue_total - (BaseCost_total + Irr_totalCost)
  
  # ---- 6) WQ prediction row ----------------------------------------
  pred_row <- tibble::tibble(
    Year = Y,
    `LandCover_m2_all cult`      = Cult_m2,
    LandCover_m2_grassPastureHay = Grass_m2,
    LandCover_m2_developed       = LC_dev_base
  ) %>%
    dplyr::left_join(
      wq_context %>%
        dplyr::select(Year, Climate_precip_m, Climate_Tmean_C, precip_percentileChange) %>%
        dplyr::distinct(Year, .keep_all = TRUE),
      by = "Year"
    ) %>%
    dplyr::mutate(
      Management_irrigation_m       = ifelse(Cult_m2 > 0, Irr_m3_total   / Cult_m2, 0),
      Management_FertilizerUse_kgm2 = ifelse(Cult_m2 > 0, Fert_kg_total / Cult_m2, 0)
    )
  
  if (anyNA(pred_row)) return(c(PEN, PEN, PEN))
  
  pred_log1p   <- as.numeric(predict(wq_model_log1p, newdata = pred_row))
  nitrate_kgyr <- max(0, exp(pred_log1p) - 1)
  
  c(nitrate_kgyr, -NR_total, Irr_m3_total)
}


```

 - Aggregate across ALL years: mean N flux (kg/yr) and mean -profit ($/yr)
  - this fx repeats all of ^^ fx but then takes the average of all the years, so a basin-wide, period-wide average of NR, WQ, irr for a specific land use configuration
```{r}
eval_candidate_all_years <- function(
  x,
  years_vec,
  df_opt,
  wq_model_log1p,
  wq_context,
  lu_baseline,
  universal_costs,
  crop_params,
  baseline_irrig_frac,
  hist_mix = NULL,
  weights = NULL,
  na_penalty = 1e9
){
  stopifnot(length(years_vec) > 0)
  
  vals <- vapply(
    years_vec,
    \(Y) eval_candidate(
      x                = x,
      Y                = Y,
      df_opt           = df_opt,
      wq_model_log1p   = wq_model_log1p,
      wq_context       = wq_context,
      lu_baseline      = lu_baseline,
      universal_costs  = universal_costs,
      crop_params      = crop_params,
      baseline_irrig_frac = baseline_irrig_frac,
      hist_mix         = hist_mix
    ),
    numeric(3)
  )
  vals <- t(vals)
  colnames(vals) <- c("n_flux_kgyr", "neg_profit_usdyr", "irr_m3_yr")
  
  # weights (default: equal for finite rows)
  if (is.null(weights)) weights <- rep(1, nrow(vals))
  if (length(weights) != nrow(vals)) stop("weights must match length(years_vec).")
  
  wmean <- function(v, w) {
    keep <- is.finite(v) & is.finite(w) & w > 0
    if (!any(keep)) return(NA_real_)
    wk <- w[keep] / sum(w[keep])
    sum(v[keep] * wk)
  }
  
  mean_n_flux    <- wmean(vals[, "n_flux_kgyr"],       weights)
  mean_neg_prof  <- wmean(vals[, "neg_profit_usdyr"],  weights)
  mean_irr_m3_yr <- wmean(vals[, "irr_m3_yr"],         weights)
  
  if (!is.finite(mean_n_flux) || !is.finite(mean_neg_prof) || !is.finite(mean_irr_m3_yr)) {
    return(c(na_penalty, na_penalty, na_penalty))
  }
  
  c(mean_n_flux, mean_neg_prof, mean_irr_m3_yr)
}

```
 
 - NSGA-II for the robust (mean) frontier

```{r}
optimize_all_years_nsga2 <- function(
  years_vec,
  pop  = 20,
  gens = 20,
  baseline_irrig_frac = 0.48,   # fixed historical irrigated fraction
  seed = 123
){
  stopifnot(length(years_vec) > 0)
  set.seed(seed)

  # Objective wrapper: calls your simplified eval_candidate_all_years
  fn <- function(x) {
    eval_candidate_all_years(
      x              = x,
      years_vec      = years_vec,
      df_opt         = df_opt,
      wq_model_log1p = wq_model_log1p,
      wq_context     = wq_context,
      lu_baseline    = lu_baseline,
      universal_costs = universal_costs,
      crop_params    = crop_params,
      baseline_irrig_frac = baseline_irrig_frac,
            hist_mix       = NULL
    )
  }

  # Decision vector is now only 3D: [Corn, Soybeans, Sorghum]
  # Bounds taken from historical ymin/ymax
  lower <- c(
    Corn     = crop_min["Corn"],
    Soybeans = crop_min["Soybeans"],
    Sorghum  = crop_min["Sorghum"]
  )

  upper <- c(
    Corn     = crop_max["Corn"],
    Soybeans = crop_max["Soybeans"],
    Sorghum  = crop_max["Sorghum"]
  )

# Constraints: keep all four crops within their historical ranges
  # Wheat is implied: Wheat = 1 - (Corn + Soybeans + Sorghum)
  cons <- function(x) {
    corn <- x[1]
    sorg  <- x[2]
    soy <- x[3]
    wheat <- 1 - corn - soy - sorg

    c(
      # Corn within [min, max]
      corn  - crop_min["Corn"],      # >= 0  => corn >= min
      crop_max["Corn"]     - corn,   # >= 0  => corn <= max

      # Soy within [min, max]
      soy   - crop_min["Soybeans"],
      crop_max["Soybeans"] - soy,

      # Sorghum within [min, max]
      sorg  - crop_min["Sorghum"],
      crop_max["Sorghum"]  - sorg,

      # Implied Wheat within [min, max]
      wheat - crop_min["Wheat"],
      crop_max["Wheat"] - wheat
    )
  }


  res <- mco::nsga2(
    fn,
    idim         = 3,   # 3 decision variables
    odim         = 3,   # N flux, -profit, irrigation
    lower.bounds = lower,
    upper.bounds = upper,
    popsize      = pop,
    generations  = gens,
    constraints  = cons,
    cdim         = 8
  )

  out <- as.data.frame(res$par)
  names(out) <- c("Corn", "Soybeans", "Sorghum")
  out$Wheat  <- 1 - out$Corn - out$Soybeans - out$Sorghum

  out$Nitrate_obj  <- res$value[, 1]   # mean nitrate (kg/yr) – minimize
  out$NR_obj       <- -res$value[, 2]  # mean net returns ($/yr) – maximize
  out$Irrig_m3_obj <- res$value[, 3]   # mean irrigation volume (m3/yr) – minimize

  out
}

```
 
 
 - runing alg & saving output
```{r}
years_all <- sort(unique(df_opt$Year))

start_time <- Sys.time()
robust_out <- optimize_all_years_nsga2(
  years_all, pop = 20, gens = 20,
  seed = 42,
  baseline_irrig_frac = 0.48   # or your historical fraction

)

end_time <- Sys.time()

execution_time <- end_time - start_time
print(execution_time)

timestamp <- format(Sys.time(), "%Y%m%d_%H%M")
saveRDS(robust_out, file = paste0("robust_out_baseline_test_short", timestamp, ".rds"))

```
 
 
 - Frontier filter (on aggregated objectives)
```{r}
#robust_frontier <- robust_out %>%
 # arrange(Nitrate_obj, desc(NRha_obj), ) %>%
#  mutate(cum_best = cummax(NRha_obj)) %>%
#  filter(NRha_obj >= cum_best) %>%
 # select(-cum_best)


pareto_nd <- function(df, cols, minimize, tol = 0) {
  M <- as.matrix(df[, cols, drop = FALSE])
  # Convert "maximize" to "minimize" by negating that column
  for (k in seq_along(cols)) if (!minimize[k]) M[, k] <- -M[, k]
  n <- nrow(M); keep <- rep(TRUE, n)
  for (i in seq_len(n)) {
    if (!keep[i]) next
    # j dominates i if j is <= i in all cols (with tol) and < in at least one
    less_equal_all <- sweep(M, 2, M[i, ], `<=`) | abs(sweep(M, 2, M[i, ], `-`)) <= tol
    strictly_less  <- sweep(M, 2, M[i, ], `<`)  & !abs(sweep(M, 2, M[i, ], `-`)) <= tol
    dominates_i <- apply(less_equal_all, 1, all) & apply(strictly_less, 1, any)
    dominates_i[i] <- FALSE
    if (any(dominates_i)) keep[i] <- FALSE
  }
  keep
}

cols3     <- c("Nitrate_obj", "NRha_obj", "Irrig_m3_obj")
minimize3 <- c(TRUE, FALSE, TRUE)  # min N, max $, min Irr
keep <- pareto_nd(robust_out, cols3, minimize3, tol = 0)

robust_frontier <- robust_out[keep, ] %>%
  arrange(Nitrate_obj, desc(NRha_obj), Irrig_m3_obj)

#saveRDS(robust_frontier, file = paste0("robust_out_frontier_diversified_short", timestamp, ".rds"))

```
 
 
 
 - viz
```{r}
ggplot(robust_out, aes(x = Nitrate_obj, y = NR_obj, color = Irrig_m3_obj)) +
  geom_point(size = 3.5, alpha = 0.8) +
  scale_color_viridis_c(direction = -1,  trans = "log10") +
  theme_test(base_size = 14) +
  labs(
    x = "Nitrate flux (kg/yr, ↓ better)",
    y = "Net Return ($/yr, ↑ better)",
    color = "Irrigation (m³/yr, ↓ better)",
    title = "Baseline: Nitrate vs Profit (colored by irrigation)"
  ) + theme(plot.title.position = "plot")

library(plotly)

# Make an interactive 3D scatter
plot_ly(
  data = robust_out,
  x = ~Nitrate_obj,  # minimize
  y = ~NR_obj,   # maximize
  z = ~Irrig_m3_obj,    # minimize
  color = ~Corn,
  colors = viridisLite::viridis(100),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 4, opacity = 0.7)
) %>%
  layout(
    scene = list(
      xaxis = list(title = "Nitrate (scaled, ↓ better)"),
      yaxis = list(title = "Net Return (scaled, ↑ better)"),
      zaxis = list(title = "Irrigation Use (scaled, ↓ better)")
    ),
    title = "3D Pareto Front: Nitrate vs Profit vs Irrigation"
  )


#install.packages("scatterplot3d")
library(scatterplot3d)

scatterplot3d(
  x = robust_out$Nitrate_obj,
  y = robust_out$NR_obj,
  z = robust_out$Irrig_m3_obj,
  color = "darkblue",
  pch = 19,
  main = "3D Pareto Front (scaled values)",
  xlab = "Nitrate (↓)",
  ylab = "Profit (↑)",
  zlab = "Irrigation (↓)"
)


```
 
 # diagnositc tables
```{r}
robust_out %>%
  summarise(
    min_corn = min(Corn), max_corn = max(Corn),
    min_soy  = min(Soybeans), max_soy = max(Soybeans),
    min_sorg = min(Sorghum), max_sorg = max(Sorghum),
    min_wht  = min(Wheat),   max_wht = max(Wheat)
  )


percrop_summary <- df_opt %>%
  dplyr::group_by(Crop) %>%
  dplyr::summarise(
    mean_yield_kg_ha = mean(Yield_kgHa, na.rm = TRUE),
    mean_irr_m       = mean(Irrigation_m, na.rm = TRUE),
    mean_crop_ha  = mean(Crop_area_ha, na.rm = TRUE)  

  ) %>%
  dplyr::left_join(
    crop_params %>% 
      dplyr::select(Crop, income_per_kg, total_cost_per_kg, net_return_per_kg, fert_kgha ),
    by = "Crop"
  ) %>%
  dplyr::mutate(
    revenue_per_ha    = mean_yield_kg_ha * income_per_kg,
    cost_per_ha       = mean_yield_kg_ha * total_cost_per_kg,
    net_return_per_ha = mean_yield_kg_ha * net_return_per_kg,
    fert_kg = mean_crop_ha * fert_kgha
    
  )

percrop_summary


df_opt %>%
  group_by(Crop) %>%
  summarise(
    mean_Irr_m_dfopt = mean(Irrigation_m, na.rm = TRUE)
  ) %>%
  left_join(crop_params %>% select(Crop, irr_depth_m), by = "Crop")

```
 
 
 
 